---
title: "Analysis of Severe Weather Events Impacts"
author: "jnabonne"
date: "7/17/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=F, message=F)
```

---

## About
This document computes the answer to the [assignment #2 of Reproducible Research course](https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2).

### Dataset and corresponding Documentation
The dataset for this exercice available 
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

There is also two documents to better understand its content:

* [NWS Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* [additional FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


### Synopsis of the Analysis

The main objective of this analysis is to understand the possible impacts 
(economic and public health) of server weather events.  
Two main questions were asked : 

1. which types of events are most harmful to population health?
2. which types of events have the greatest economic consequences?

This analysis starts by downloading a dataset from the U.S. National Oceanic 
and Atmospheric Administration's (NOAA) storm database containing 
characteristics of major storms and weather events in the US 
(time, location, impacts estimation).

The data are then cleaned to only conserve interesting measures and in adequate 
format for manipulation and display.

Some new tables presenting aggregated results are the created giving for each 
type of events their number of occurences, victims and damages amount.

The Analysis ends by displaying some graphics showing evidence that tornados 
and flood are respectively the worst events in term of human victims 
and economical damages.


---

## Data Processing

First we load the libraries used for this analysis
```{r loading_libs, message=F, warning=FALSE}
    library(ggplot2)
    library(dplyr)
```

Then we download the dataset from the internet.  
No check on the already existing presence of the file is done, this to ensure 
that we always work on the latest available version of the dataset.

```{r data_downloading, cache=TRUE}
	# DOWNLOADING DATASET (EACH TIME TO GET ITS LATEST VERSION)
	file_url  <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
	file_path <- "StormData.csv.bz2"
	download.file(file_url, destfile=file_path, method="curl")
```

The dataset is then loaded
```{r data_loading, cache=TRUE}
	# Reading (and unzipping on-the-fly) the DATASET
	sdata <- read.csv("StormData.csv.bz2")
#	sdata <- read.csv("StormData.csv.bz2", nrows=30000) # only to speed tests ;)
```

Then some processing are done to the dataset to ease the rest of the analysis.  

First we change the exponent factors to be applied to property and crop damages 
by correct numeric values. Following some reading on the internet 
(especially [this post](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html))
the following matching as been performed

|   symbol   | exponent value |
|:----------:|---------------:|
| numeric    |             10 |
| h or H     |            100 |
| t or T     |           1000 |
| k or K     |           1000 |
| m or M     |      1.000.000 |
| b or B     |  1.000.000.000 |
| +          |              1 |
| all others |              0 |

These exponent values are then used to redress property and crop damages.

```{r desactivated_code, warning=F, message=F, echo=F}
# DESACTIVATED CODE to manage exponential facors (if you have any idea why...)
#   prop_exp <- data.frame(exp=unique(sdata$PROPDMGEXP))
#   crop_exp <- data.frame(exp=unique(sdata$CROPDMGEXP))
#   eval_exp <- data.frame(exp=unique(rbind(prop_exp, crop_exp)))
#   eval_exp <- mutate(eval_exp, value=ifelse(exp %in% c('h','H'), 100,
#                                      ifelse(exp %in% c('t','T','k','K'), 1000,
#                                      ifelse(exp %in% c('m','M'), 1000000,
#                                      ifelse(exp %in% c('b','B'), 1000000000,
#                                      ifelse(exp %in% 0:9, 10,
#                                      ifelse(exp == '+', 1, 0)))))))
#   sdata <- mutate(sdata, 
#                   PROPDMG= PROPDMG * prop_exp$value[prop_exp$exp==PROPDMGEXP],
#                   CROPDMG= CROPDMG * crop_exp$value[crop_exp$exp==CROPDMGEXP])
```

```{r data_cleaning_1, cache=TRUE}
    # Mathcing exponantial factors with correct numeric values
    sdata <- mutate(sdata, CROPDMGEXP=ifelse(CROPDMGEXP %in% c('h','H'), 100,
                            ifelse(CROPDMGEXP %in% c('t','T','k','K'), 1000,
                            ifelse(CROPDMGEXP %in% c('m','M'), 1000000,
                            ifelse(CROPDMGEXP %in% c('b','B'), 1000000000,
                            ifelse(CROPDMGEXP %in% 0:9, 10,
                            ifelse(CROPDMGEXP == '+', 1, 0)))))))
    sdata <- mutate(sdata, PROPDMGEXP=ifelse(PROPDMGEXP %in% c('h','H'), 100,
                            ifelse(PROPDMGEXP %in% c('t','T','k','K'), 1000,
                            ifelse(PROPDMGEXP %in% c('m','M'), 1000000,
                            ifelse(PROPDMGEXP %in% c('b','B'), 1000000000,
                            ifelse(PROPDMGEXP %in% 0:9, 10,
                            ifelse(PROPDMGEXP == '+', 1, 0)))))))
    # Computing final damages amount from corresponding exponential factors
    sdata <- mutate(sdata,PROPDMG=PROPDMG*PROPDMGEXP,CROPDMG=CROPDMG*CROPDMGEXP)
```

The dataset is the cleaned by removing unusefull columns.  
The dates, even if not very used in this analyis (kept to allow playing with 
the data set in eda mode) are casted correctly to be more easily manageable.

```{r data_cleaning_2}
    # Keeping only interresting data
    sdata <- subset(sdata, select=c("EVTYPE",   "BGN_DATE", "FATALITIES", 
                                    "INJURIES", "PROPDMG",  "CROPDMG"))

    # Casting dates correctly (date is here optional and was kept for tests)
    sdata$BGN_DATE <- as.Date(sdata$BGN_DATE, "%m/%d/%Y %H:%M:%S")
    #sd1 <- sdata %>% group_by(year(BGN_DATE), EVTYPE) # requires lubridate lib
```

Finally, the dataset is grouped by event type (using the `dplyr` library) and 
summarized creating totals for the interesting measures:

* nb of occurences
* total victims, with subtotal per injuries and fatalities
* total damages, with subtotal for property and crop damages

To get plots that can easily be read by humans, we only kept the top10 events 
(in term of occurences, victims and damages) otherwise we would get graphics 
with hundreds of variables

```{r data_summarizing}
    # Grouping data by events and summarizing
    sd2 <- sdata %>% group_by(EVTYPE)
    sd3 <- summarize(sd2, total_events=length(EVTYPE), 
                     total_injuries=sum(INJURIES),
                     total_fatalities=sum(FATALITIES),
                     total_victims=sum(INJURIES)+sum(FATALITIES),
                     total_property_damages=sum(PROPDMG),
                     total_crop_damages=sum(CROPDMG),
                     total_damages=sum(PROPDMG)+sum(CROPDMG))
    
    # Building top10 to get cleaner plots
    top10_events  <- arrange(sd3, desc(sd3$total_events))[1:10,]
    top10_victims <- arrange(sd3, desc(sd3$total_victims))[1:10,]
    top10_damages <- arrange(sd3, desc(sd3$total_damages))[1:10,]
```


---

## Results

From a pure events based approach we can see on the below graphics **the type of
event that occurs most is `r as.character(top10_events$EVTYPE[1])`.**  
The graphic below shows the most frequent climatic events in the US between 
1950 and 2011:
```{r plotting_events}
    # ploting nb occurrences / events type
    g <- ggplot(data=top10_events, 
            aes(x=reorder(EVTYPE, -total_events), y=total_events, fill=EVTYPE))
    # printing plot with adequate legends and scaling
    g + geom_bar(stat="identity") + theme_minimal() +
    ggtitle("NUMBER OF OCCURENCES PER EVENTS TYPE") + labs(fill="Events") +
    scale_y_continuous(name="number of occurences", labels = scales::comma) +
    theme(axis.title.x=element_blank(), axis.text.x=element_blank(), 
                                                axis.ticks.x=element_blank())
```

This does not represent a list of the most dangerous climatic events. We should 
take some criteria to investigate more about these catastrophic events...

### Victims based approach

But if we base our analysis on the number of victims the result is quite 
different; **the worst event in terms of human victims 
is `r as.character(top10_victims$EVTYPE[1])`.**  
Note that have been summed here injuries and fatalities (even if deaths could 
arguably count more) as it is difficult to create a scale between these two.  

The table below displays a top10 of victims per event type:
```{r tabling_victims}
    subset(arrange(sd3, desc(sd3$total_victims))[1:10,], select=c("EVTYPE", 
        "total_events", "total_injuries", "total_fatalities", "total_victims"))
```

The graphic below shows a top10 of the worst events in terms of victims:
```{r plotting_victims} 
    # ploting nb victims / events type
    g <- ggplot(data=top10_victims, 
        aes(x=reorder(EVTYPE, -total_victims), y=total_victims, fill=EVTYPE))
    # printing plot with adequate legends and scaling
    g + geom_bar(stat="identity") + theme_minimal() +
    ggtitle("TOTAL NUMBER OF VICTIMS PER EVENTS TYPE") + labs(fill="Events") +
    scale_y_continuous(name="number of injuries + fatalities", 
                                                    labels = scales::comma) +
    theme(axis.title.x=element_blank(), axis.text.x=element_blank(),
                                                axis.ticks.x=element_blank())
```

### Economic based approach
Now from an economic point of view, the result is again different 
with **`r as.character(top10_damages$EVTYPE[1])` arriving first 
in term of damages amount.**  
As for victims, both property and crop damages have been added together 
to get these results.  

The table below displays a top10 of economic damages per event type:
```{r tabling_damages}
    subset(arrange(sd3, desc(sd3$total_damages))[1:10,], select=c("EVTYPE", 
                                    "total_events", "total_property_damages", 
                                    "total_crop_damages", "total_damages"))
```

The graphic below shows the 10 worst events in terms of economic damages.
```{r plotting_damages} 
    # ploting damages / events type
    g <- ggplot(data=top10_damages, 
        aes(x=reorder(EVTYPE, -total_damages), y=total_damages, fill=EVTYPE))
    # printing plot with adequate legends and scaling
    g + geom_bar(stat="identity") + theme_minimal() +
    ggtitle("TOTAL AMOUNT OF DAMAGES PER EVENTS TYPE") + labs(fill="Events") +
    scale_y_continuous(name="amount of damages (property + crop) in USD", 
                                                    labels = scales::comma) +
    theme(axis.title.x=element_blank(), axis.text.x=element_blank(), 
                                                axis.ticks.x=element_blank())
```

---

This is the end, thanks for reading until it!